
---
title: "SenOlinkReport3"
author: "James Billingsley"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
---

```{r setup, include=FALSE, message=FALSE, warning = FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  error = TRUE,
  fig.align = "center",
  fig.path = "/Users/jmb714/Desktop/figures/",
  out.width = "75%",
  message = FALSE,
  warning = FALSE
)

clientname <- "Pritha Sen"
clientemail <- "PSEN@BWH.HARVARD.EDU"
labPI <- "Pritha Sen"
lablocation <- "BWH"

analystname <- "James Billingsley"
analystemail <- "jbillingsley@hsph.harvard.edu"
set.seed(42)
datadir <- data_dir <- paste0(
  "/Users/jmb714/Harvard University Dropbox/",
  "HBC Team Folder (1)/Consults/pritha_sen/",
  "sen_olink_human_blood_mpox_hbc05277/data/"
)
```


**Olink REVEAL analysis of mpox samples [hbc05277] `r clientname`. **


Contact `r analystname` (`r analystemail`) for additional details.

The most recent update of this html document occurred: `r date()`


The raw data are here: 

```{r datadir}
cat(datadir)
```

<br/>
<br/>
<br/>
<br/>

This report examines differential expression/abundance of serum proteins in the JYNNEOS Mpox prime-boost vaccine experiment. We have serial timepoints representing baseline (prime), boost, and two followup timepoints, i.e. D0, D28, D56 and D180.

We will use linear mixed effects modeling and control for donor effects.

We can look for proteins that change significantly over time, and also do post-tests to compare specific binary contrasts.

In the QC process (slightly modified here from SenOlinkRept1.Rmd), we remove any assays that have ≥ 95% of their data points below LOD, and we remove the two QC flagged assays.

This removes 79 assays leaving 953.

We also run the analysis without removing these high LOD assays for comparison.




```{r libraries}
library(OlinkAnalyze)
library(knitr)
library(patchwork)
library(tidyverse)
library(readxl)
library(UpSetR)
library(ComplexHeatmap)
library(forcats)
```


```{r readin}
MPXV <- read.table(file = "~/Desktop/MPXV_June2025_Extended_NPX_2025-06-17.csv", sep = ";", h = TRUE)

# add LOD data

MPXV <- olink_lod(MPXV, lod_file_path = "~/Desktop/Reveal_Fixed_LOD_csv_file.csv", lod_method = "FixedLOD")


# remove Control samples

no_ctrl_MPXV <- MPXV %>%
  filter(!str_detect(SampleID, regex("Control"))) # remove control samples


# remove control assays


no_ctrl_MPXV <- no_ctrl_MPXV %>%
  filter(!Assay %in% c(
    "Amplification control",
    "Extension control",
    "Incubation control"
  )) # remove control assays!


# remove flagged assays n=2


warn_assays <- unique(no_ctrl_MPXV %>% filter(AssayQC == "WARN") %>% pull(Assay)) # two assays, "NPM1"   "SDHAF4"


no_ctrl_MPXV <- no_ctrl_MPXV %>%
  filter(!Assay %in% warn_assays)

# length(unique(no_ctrl_MPXV$Assay))#1032


# remove assays with ≥ 95% of datapoints reading below their published LOD.

assay_qc_summary <- no_ctrl_MPXV %>%
  group_by(Assay) %>%
  summarise(
    pct_below_LOD = mean(NPX <= LOD)
  )


low_detect_assays <- assay_qc_summary %>%
  filter(pct_below_LOD >= 0.95) %>%
  pull(Assay)


clean_MPXV <- no_ctrl_MPXV %>%
  filter(!Assay %in% low_detect_assays) # remove high LOD assays


NatAnn <- read_excel("~/Desktop/250624 Annotations for MPXJYN Olink.xlsx")


### Fix typo

# which(NatAnn$sample_id == "JYN14-4")#62


NatAnn$sample_id[62] <- "JYN14-5"


header_Fulldf <- clean_MPXV %>%
  distinct(SampleID) %>%
  arrange(SampleID) %>%
  mutate(
    Donor = str_sub(SampleID, 1, -3)
  ) %>%
  left_join(
    NatAnn %>%
      select(sample_id, `NS Annotation`) %>%
      rename(NS_annotation = `NS Annotation`),
    by = c("SampleID" = "sample_id")
  )



Expt <- c(rep("JYN", times = 45), rep("MPX", times = 41))

header_Fulldf$Expt <- Expt


header_Fulldf$Donor <- factor(header_Fulldf$Donor)

header_Fulldf$NS_annotation <- factor(header_Fulldf$NS_annotation)

header_Fulldf$NExpt <- factor(header_Fulldf$Expt)


annotated_long <- clean_MPXV %>%
  left_join(header_Fulldf, by = "SampleID") # add annotation to my long form dataframe


JYN_long <- annotated_long %>%
  filter(Expt == "JYN")


JYN_long %>%
  select(SampleID, NS_annotation) %>%
  distinct()


# relevel


JYN_long <- JYN_long %>%
  mutate(
    NS_annotation = fct_relevel(
      NS_annotation,
      "D0", "D28", "D56", "D180",
      after = Inf
    )
  )

JYN_lmer_results <- olink_lmer(
  df       = JYN_long,
  variable = "NS_annotation",
  random   = "Donor"
)


# write.table(JYN_lmer_results, file="~/Desktop/JYN_lmer_results.txt", sep="\t", col.names=NA)


JYN_posthoc <- olink_lmer_posthoc(
  df            = JYN_long,
  variable      = "NS_annotation",
  random        = "Donor",
  effect        = "NS_annotation",
  verbose       = TRUE
)

# write.table(JYN_posthoc, file="~/Desktop/JYN_lmer_results_posthoc.txt", sep="\t", col.names=NA)


signif_counts <- JYN_posthoc %>%
  filter(Threshold == "Significant") %>%
  group_by(contrast) %>%
  summarise(n_signif = n(), .groups = "drop")


ggplot(
  signif_counts %>%
    mutate(contrast = fct_reorder(contrast, n_signif, .desc = TRUE)),
  aes(x = contrast, y = n_signif)
) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  theme_minimal() +
  labs(
    x     = "Pairwise Contrast",
    y     = "Number of Significant Proteins",
    title = "DEPs per Contrast from NS_annotation"
  )


sig_sets <- JYN_posthoc %>%
  filter(Threshold == "Significant") %>%
  distinct(Assay, contrast) %>%
  group_by(contrast) %>%
  reframe(proteins = list(Assay))



sig_list <- setNames(sig_sets$proteins, sig_sets$contrast)


upset(fromList(sig_list),
  order.by = "freq",
  mb.ratio = c(0.6, 0.4),
  main.bar.color = "steelblue",
  sets.bar.color = "skyblue",
  nsets = 6,
  nintersects = NA
)



sig_proteins_lmm <- JYN_lmer_results %>%
  filter(term == "NS_annotation", Threshold == "Significant") %>%
  pull(Assay)


sig_npx <- JYN_long %>%
  filter(Assay %in% sig_proteins_lmm)

mat <- sig_npx %>%
  select(SampleID, Assay, NPX) %>%
  pivot_wider(names_from = SampleID, values_from = NPX)

rownames(mat) <- mat$Assay

mat <- as.matrix(mat[, -1])

mat_scaled <- t(scale(t(mat)))


my_pal <- colorRampPalette(c("blue", "white", "red"))(256)


Heatmap(
  mat_scaled,
  name = "Z-score",
  cluster_rows = TRUE,
  cluster_columns = TRUE,
  col = my_pal,
  show_row_names = TRUE,
  show_column_names = TRUE,
  column_title = "Samples",
  row_title = "Significant Proteins"
)
```
<br/>
<br/>

Comparing significantly differentially expressed proteins between D0 and D56, there are 168 proteins significantly downregulated at D56 and 21 proteins significantly upregulated at D56.

Significance is at an adjusted p value of < 0.05.

In the output file "estimate" represents log2FC.

<br/>
<br/>
<br/>
<br/>

Some example boxplots


```{r plotsB}
baseline_vs_t3 <- JYN_posthoc %>%
  filter(str_detect(contrast, "D0 - D56")) %>%
  filter(Threshold == "Significant") %>%
  arrange(Adjusted_pval)


# baseline_vs_t3 %>% filter(estimate > 0) %>% nrow()#168

# baseline_vs_t3 %>% filter(estimate < 0) %>% nrow()#21


# head(baseline_vs_t3 %>% filter(estimate > 0))

# baseline_vs_t3 %>% filter(estimate < 0)


example_genes <- c("TNFRSF9", "VEGFD", "TRIM58")

example_genes2 <- c("AGRP", "CCK", "BMP6")


plot_data <- JYN_long %>%
  filter(Assay %in% example_genes)

plot_data2 <- JYN_long %>%
  filter(Assay %in% example_genes2)




ggplot(plot_data, aes(x = NS_annotation, y = NPX, fill = NS_annotation)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.2, alpha = 0.6, size = 1.5) +
  facet_wrap(~Assay, scales = "free_y") +
  theme_bw() +
  labs(
    title = "Expression of Selected Proteins by NS_annotation",
    x = "NS_annotation",
    y = "NPX"
  ) +
  scale_fill_brewer(palette = "Set2")

ggplot(plot_data2, aes(x = NS_annotation, y = NPX, fill = NS_annotation)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.2, alpha = 0.6, size = 1.5) +
  facet_wrap(~Assay, scales = "free_y") +
  theme_bw() +
  labs(
    title = "Expression of Selected Proteins by NS_annotation",
    x = "NS_annotation",
    y = "NPX"
  ) +
  scale_fill_brewer(palette = "Set2")
```
<br/>
<br/>
<br/>
<br/>


If we run the experiment with no LOD assays removed, number assays = 1032


Comparing significantly differentially expressed proteins between D0 and D56, there are 173 proteins significantly downregulated at D56 and 22 proteins significantly upregulated at D56.


```{r noLODfilter, eval=F}
# Run without removing high LOD assay_qc_summary


no_ctrl_MPXV <- MPXV %>%
  filter(!str_detect(SampleID, regex("Control"))) # remove control samples

# remove control assays


no_ctrl_MPXV <- no_ctrl_MPXV %>%
  filter(!Assay %in% c(
    "Amplification control",
    "Extension control",
    "Incubation control"
  )) # remove control assays!


# remove flagged assays n=2


warn_assays <- unique(no_ctrl_MPXV %>%
  filter(AssayQC == "WARN") %>%
  pull(Assay)) # two assays, "NPM1"   "SDHAF4"


no_ctrl_MPXV <- no_ctrl_MPXV %>%
  filter(!Assay %in% warn_assays)

clean_MPXV <- no_ctrl_MPXV

annotated_long <- clean_MPXV %>%
  left_join(header_Fulldf, by = "SampleID") # add annotation to my long form dataframe


JYN_long <- annotated_long %>%
  filter(Expt == "JYN")


JYN_long %>%
  select(SampleID, NS_annotation) %>%
  distinct()


# relevel



JYN_long <- JYN_long %>%
  mutate(
    NS_annotation = fct_relevel(
      NS_annotation,
      "D0", "D28", "D56", "D180",
      after = Inf
    )
  )


JYN_lmer_results <- olink_lmer(
  df       = JYN_long,
  variable = "NS_annotation",
  random   = "Donor"
)


# write.table(JYN_lmer_results, file="~/Desktop/JYN_lmer_results_noLODfilter.txt", sep="\t", col.names=NA)


JYN_posthoc <- olink_lmer_posthoc(
  df            = JYN_long,
  variable      = "NS_annotation",
  random        = "Donor",
  effect        = "NS_annotation",
  verbose       = TRUE
)

# write.table(JYN_posthoc, file="~/Desktop/JYN_lmer_results_posthoc_noLODfilter.txt", sep="\t", col.names=NA)


baseline_vs_t3 <- JYN_posthoc %>%
  filter(str_detect(contrast, "D0 - D56")) %>%
  filter(Threshold == "Significant") %>%
  arrange(Adjusted_pval)


baseline_vs_t3 %>%
  filter(estimate > 0) %>%
  nrow() # 173

baseline_vs_t3 %>%
  filter(estimate < 0) %>%
  nrow() # 22
```

```{r sessionInfo}
sessionInfo()
```
