
---
title: "SenOlinkReport1"
author: "James Billingsley"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
---

```{r setup, include=FALSE, message=FALSE, warning = FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  error = TRUE,
  fig.align = "center",
  fig.path = "/Users/jmb714/Desktop/figures/",
  out.width = "75%",
  message = FALSE,
  warning = FALSE
)

clientname <- "Pritha Sen"
clientemail <- "PSEN@BWH.HARVARD.EDU"
labPI <- "Pritha Sen"
lablocation <- "BWH"

analystname <- "James Billingsley"
analystemail <- "jbillingsley@hsph.harvard.edu"
set.seed(42)
datadir <- data_dir <- paste0(
  "/Users/jmb714/Harvard University Dropbox/",
  "HBC Team Folder (1)/Consults/pritha_sen/",
  "sen_olink_human_blood_mpox_hbc05277/data/"
)
```

**Olink REVEAL analysis of mpox samples [hbc05277] `r clientname`. **


Contact `r analystname` (`r analystemail`) for additional details.

The most recent update of this html document occurred: `r date()`


The raw data are here: 

```{r datadir}
cat(datadir)
```

<br/>
<br/>
<br/>
<br/>

Olink Reveal data; Examining serum proteins in:

1. Time course analysis of donors infected with Mpox, acute and resolution phases. 18 donors, Most with 3 timepoints, one with 4, some with a single timepoint.

2. Time course analysis of participants receiving JYNNEOS mpox/smallpox prime-boost vaccine. Three timepoints, baseline (dose1), boost(dose2), postvacc.


We have a single plate, 96 samples and 1037 assays. Some of the 96 samples are plate control samples and some of the assays are control assays.


We'll examine each of the two experiments individually, and combined.

Olink data (NPX) generally comes normalized with assay and sample QC flags from Olink. The experiment here uses the Reveal technology which measures approx 1000 proteins per well. 

This experiment is a single plate experiment so no further normalization was required. If you have a multiplate experiment, additional normalization ,such as "bridging normalization" (between plates) will likely be required. This is readily done using the Olink_analyze R package. The package also provides convenient qc plotting functions and differential expression functions, both of which I use in this analysis.

For QC, I look for flagged samples and assays, (no sample flags, 2 assay flags). I also do PCA plots looking for samples outliers (none found), and look for poorly performing assays. 

To check for poorly performing assays, I look, for each assay, at the number of data points that fall below the Limit Of Detection.  In a multiplate experiment you can calculate the LOD for each assay using Sample controls on each plate. But for a single plate assay you cannot. In that case such as ours, Olink provides an LOD file that can be used as a substitute. 

I plotted the percentage of below LOD data points for each assay. There was no obvious inflection point on the plot, so I chose to exclude any assays with greater than 95% below LOD data points. I think this is a sensible threshold for this technology and experiment. As a sanity check I also ran differential expression with no assays removed, and the results were quite similar. 

Note, in this report I did not remove the two qc flagged assays. Please see the differential expression analyses reports for this project, for examples of filtering flagged assays.


https://github.com/Olink-Proteomics/OlinkRPackage

https://cran.r-project.org/web/packages/OlinkAnalyze/vignettes/LOD.html



```{r libraries}
library(tidyverse)
library(Matrix)
library(RCurl)
library(knitr)
library(patchwork)
library(gridExtra)
library(reshape2)
library(Matrix.utils)
library(DESeq2)
library(EnhancedVolcano)
library(OlinkAnalyze)
library(readxl)
```


```{r readin}
mpxv <- read.table(
  file = "~/Desktop/MPXV_June2025_Extended_NPX_2025-06-17.csv",
  sep  = ";",
  h    = TRUE
)

glimpse(mpxv)
hist(mpxv$NPX)

mpxv %>%
  summarise(
    n_samples   = n_distinct(SampleID),
    n_olink_ids = n_distinct(OlinkID),
    n_assays    = n_distinct(Assay),
    n_panels    = n_distinct(Panel),
    n_wells     = n_distinct(WellID),
    n_plate_ids = n_distinct(PlateID)
  )
```
<br/>
<br/>
<br/>
<br/>

Some plots of samples in the full dataset

```{r plotsA}
olink_dist_plot(mpxv, color_g = "SampleQC") +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, size = 5)
  )

olink_qc_plot(mpxv, color_g = "SampleQC", label_outliers = TRUE)

olink_pca_plot(mpxv, color_g = "SampleQC", label_samples = TRUE)

npx_flt <- mpxv %>%
  filter(!str_detect(SampleID, regex("Control")))

olink_dist_plot(npx_flt, color_g = "SampleQC") +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, size = 5)
  ) +
  ggtitle("Control samples removed")

olink_qc_plot(
  npx_flt,
  color_g         = "SampleQC",
  label_outliers  = TRUE
) +
  ggtitle("Control samples removed")

olink_pca_plot(npx_flt, color_g = "SampleQC", label_samples = TRUE)
```

<br/>
<br/>
<br/>
<br/>

Assay QC

We'll look for assays that have a high percentage of normalized counts below published LOD.


```{r qcAssays}
mpxv <- olink_lod(
  mpxv,
  lod_file_path = "~/Desktop/Reveal_Fixed_LOD_csv_file.csv",
  lod_method    = "FixedLOD"
)

assay_qc_summary <- mpxv %>%
  group_by(OlinkID, Assay, AssayQC) %>%
  summarise(
    pct_below_lod = mean(NPX <= LOD, na.rm = TRUE),
    cv            = sd(NPX, na.rm = TRUE) / mean(NPX, na.rm = TRUE),
    .groups       = "drop"
  ) %>%
  arrange(desc(pct_below_lod), desc(cv))

assay_qc_summary %>%
  filter(pct_below_lod > 0.90) %>%
  summarise(num_assays = n())

ranked_assays <- assay_qc_summary %>%
  arrange(pct_below_lod) %>%
  mutate(order = row_number())

ggplot(ranked_assays, aes(x = order, y = pct_below_lod)) +
  geom_point(color = "steelblue") +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(
    title = "Percentage Below LOD by Assay (Ranked)",
    x     = "Assay Rank (Low to High % Below LOD)",
    y     = "% of Samples Below LOD"
  ) +
  theme_minimal()
```

<br/>
<br/>
<br/>
<br/>

The QC filter cutoff we use to reduce noise, (if we use one), is somewhat arbitrary. But I'll use here 90% as an example.

70 assays have > 90 % data points below LOD

If I use a 95% threshold it removes 35 assays

If I use a 99% threshold it removes 10 assays



```{r filter_low_detect}
low_detect_assays <- assay_qc_summary %>%
  filter(pct_below_lod >= 0.90) %>%
  pull(OlinkID)

clean_mpxv <- mpxv %>%
  filter(!OlinkID %in% low_detect_assays) %>%
  filter(!str_detect(SampleID, regex("Control"))) %>%
  filter(
    !Assay %in% c(
      "Amplification control",
      "Extension control",
      "Incubation control"
    )
  )

olink_pca_plot(clean_mpxv, color_g = "SampleQC", label_samples = TRUE)
```

<br/>
<br/>
<br/>
<br/>


```{r annotateFull}
nat_ann <- read_excel("~/Desktop/250624 Annotations for MPXJYN Olink.xlsx")


header_full_df <- clean_mpxv %>%
  distinct(SampleID, WellID) %>%
  arrange(SampleID) %>%
  mutate(donor = str_sub(SampleID, 1, -3))

counts_wide_full <- clean_mpxv %>%
  select(SampleID, Assay, NPX) %>%
  pivot_wider(
    names_from   = Assay,
    values_from  = NPX,
    values_fill  = 0
  ) %>%
  arrange(SampleID)




header_full_df <- header_full_df %>%
  left_join(
    nat_ann %>% select(sample_id, `NS Annotation`),
    by = c("SampleID" = "sample_id")
  ) %>%
  dplyr::rename(ns_annotation = `NS Annotation`)

expt <- c(rep("JYN", times = 45), rep("MPX", times = 41))
header_full_df$expt <- expt


final_df_full <- header_full_df %>%
  left_join(counts_wide_full, by = "SampleID")
```
<br/>
<br/>
<br/>
<br/>

PCA MPX alone


```{r pcampx}
pca <- prcomp(
  final_df_full[c(46:86), -c(1:5)],
  scale. = TRUE
)


pca_scores <- as.data.frame(pca$x) %>%
  mutate(
    SampleID = final_df_full$SampleID[c(46:86)],
    Donor = final_df_full$donor[c(46:86)],
    NS_ann = final_df_full$ns_annotation[c(46:86)],
    Expt = final_df_full$expt[c(46:86)]
  )


ggplot(pca_scores, aes(x = PC1, y = PC2, color = NS_ann, label = SampleID)) +
  geom_point(size = 3) +
  geom_text_repel(size = 3, max.overlaps = Inf) +
  labs(
    title = "PCA of Mpx data alone",
    x = paste0("PC1 (", round(100 * summary(pca)$importance[2, 1], 1), "% variance)"),
    y = paste0("PC2 (", round(100 * summary(pca)$importance[2, 2], 1), "% variance)")
  ) +
  theme_minimal()


ggplot(pca_scores, aes(x = PC1, y = PC2, color = Donor)) +
  geom_point(size = 3) +
  labs(
    title = "PCA of Mpx data alone",
    x = paste0("PC1 (", round(100 * summary(pca)$importance[2, 1], 1), "% variance)"),
    y = paste0("PC2 (", round(100 * summary(pca)$importance[2, 2], 1), "% variance)")
  ) +
  theme_minimal()

ggplot(pca_scores, aes(x = PC1, y = PC3, color = Donor)) +
  geom_point(size = 3) +
  labs(
    title = "PCA of Mpx data alone",
    x = paste0("PC1 (", round(100 * summary(pca)$importance[2, 1], 1), "% variance)"),
    y = paste0("PC3 (", round(100 * summary(pca)$importance[2, 3], 1), "% variance)")
  ) +
  theme_minimal()
``` 
  
Very good segregation by time (and donor)
 
<br/>
<br/>
<br/>
<br/>

PCA JYNNEOS alone

```{r pca_JYN}
pca_jyn <- prcomp(
  final_df_full[1:45, -(1:5)],
  scale. = TRUE
)

pca_scores_jyn <- pca_jyn$x %>%
  as.data.frame() %>%
  mutate(
    sample_id     = final_df_full$SampleID[1:45],
    donor         = final_df_full$donor[1:45],
    ns_annotation = final_df_full$ns_annotation[1:45],
    expt          = final_df_full$expt[1:45]
  )

ggplot(
  pca_scores_jyn,
  aes(x = PC1, y = PC2, color = ns_annotation)
) +
  geom_point(size = 3) +
  labs(
    title = "PCA of JYN data alone",
    x = sprintf(
      "PC1 (%0.1f%% variance)",
      100 * summary(pca_jyn)$importance[2, 1]
    ),
    y = sprintf(
      "PC2 (%0.1f%% variance)",
      100 * summary(pca_jyn)$importance[2, 2]
    )
  ) +
  theme_minimal()

ggplot(
  pca_scores_jyn,
  aes(x = PC1, y = PC2, color = donor)
) +
  geom_point(size = 3) +
  labs(
    title = "PCA of JYN data alone",
    x = sprintf(
      "PC1 (%0.1f%% variance)",
      100 * summary(pca_jyn)$importance[2, 1]
    ),
    y = sprintf(
      "PC2 (%0.1f%% variance)",
      100 * summary(pca_jyn)$importance[2, 2]
    )
  ) +
  theme_minimal()

ggplot(
  pca_scores_jyn,
  aes(x = PC1, y = PC3, color = ns_annotation)
) +
  geom_point(size = 3) +
  labs(
    title = "PCA of JYN data alone",
    x = sprintf(
      "PC1 (%0.1f%% variance)",
      100 * summary(pca_jyn)$importance[2, 1]
    ),
    y = sprintf(
      "PC3 (%0.1f%% variance)",
      100 * summary(pca_jyn)$importance[2, 3]
    )
  ) +
  theme_minimal()

ggplot(
  pca_scores_jyn,
  aes(x = PC3, y = PC2, color = ns_annotation)
) +
  geom_point(size = 3) +
  labs(
    title = "PCA of JYN data alone",
    x = sprintf(
      "PC3 (%0.1f%% variance)",
      100 * summary(pca_jyn)$importance[2, 3]
    ),
    y = sprintf(
      "PC2 (%0.1f%% variance)",
      100 * summary(pca_jyn)$importance[2, 2]
    )
  ) +
  theme_minimal()
```


Some segregation by time, (0 is different than 56) and by donor

<br/>
<br/>
<br/>
<br/>
  

PCA all data


```{r pca_full}
pca_full <- prcomp(
  final_df_full[, -(1:5)],
  scale. = TRUE
)

pca_scores_full <- pca_full$x %>%
  as.data.frame() %>%
  mutate(
    sample_id     = final_df_full$SampleID,
    donor         = final_df_full$donor,
    ns_annotation = final_df_full$ns_annotation,
    expt          = final_df_full$expt
  )

ggplot(
  pca_scores_full,
  aes(x = PC1, y = PC2, color = expt)
) +
  geom_point(size = 3) +
  labs(
    title = "PCA of all data combined",
    x = sprintf(
      "PC1 (%0.1f%% variance)",
      100 * summary(pca_full)$importance[2, 1]
    ),
    y = sprintf(
      "PC2 (%0.1f%% variance)",
      100 * summary(pca_full)$importance[2, 2]
    )
  ) +
  theme_minimal()

ggplot(
  pca_scores_full,
  aes(x = PC1, y = PC2, color = ns_annotation)
) +
  geom_point(size = 3) +
  labs(
    title = "PCA of all data combined",
    x = sprintf(
      "PC1 (%0.1f%% variance)",
      100 * summary(pca_full)$importance[2, 1]
    ),
    y = sprintf(
      "PC2 (%0.1f%% variance)",
      100 * summary(pca_full)$importance[2, 2]
    )
  ) +
  theme_minimal()
```
<br/>
<br/>
<br/>
<br/>

The Mpx and JYN data fairly segregate, no obvious clustering with a specific cluster of Mpx data, maybe closest to the 2 timepoint?

```{r sessionInfo}
sessionInfo()
```
